\begin{abstract}
\addchaptertocentry{\abstractname} % Add the abstract to the table of contents

Click-through rate (CTR) prediction lies at the heart of the online advertising ecosystem and recommendation systems, helping to improve user engagement and platform revenue. With the advent of Pretrained Language Models (PLMs), researchers focus on incorporating textual features to enhance semantic understanding in CTR prediction. However, existing methods typically aggregate a wealth of textual features and encode the informative text into a single semantic embedding. This mechanism leads to entangled embedding that fails to capture fine-grained feature interactions, ultimately limiting CTR prediction performance. 

To address this issue, we propose Multi-faceted Semantic Disentanglement for CTR prediction (MSD-CTR), a novel framework designed to disentangle and leverage multi-faceted textual information. MSD-CTR consists of two key components: Disentangled Semantic Topic Model (DSTopic) and Topic Guided Disentangled Representation Learning (TopicDRL). DSTopic employs a disentangled generative process to extract multi-faceted knowledge from the entangled textual information. Meanwhile, TopicDRL integrates the extracted multi-faceted knowledge into CTR prediction and introduces two alignment losses to guide disentangled semantic embedding learning. 

Extensive experiments on four real-world datasets demonstrate that MSD-CTR outperforms existing CTR models, highlighting the effectiveness of disentangling textual information for better click-through rate prediction.

\end{abstract}